---
title: "Quick Start with mlr3mbspls"
author: "Stefan Coors, Clara Vetter"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start with mlr3mbspls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  dev = "svg",
  eval = TRUE,
  fig.align = "center",
  fig.height = 4.5,
  fig.path = file.path(tempdir(), "mlr3mbspls-quickstart", ""),
  fig.width = 7
)
```

# Goal

This vignette shows a straightforward, script-style MB-sPLS workflow:

- Uses the packaged synthetic `TaskMultiBlock` `mbspls_synthetic_blocks`.
- Notes the matching supervised toy tasks `mbspls_synthetic_classif` and `mbspls_synthetic_regr`.
- Generic naming only.
- Sequential objects that are easy to inspect in an interactive R session.
- Executes every supported permutation-inference route and shows its output.
- Embeds plots from temporary devices without retaining figure, result, model,
  or cache files in the working directory.

The sequence mirrors a typical analysis run:

1. Load the packaged synthetic multi-block task.
2. Run and plot the supported permutation tests.
3. Show descriptive bootstrap uncertainty.
4. Inspect task quality and define site correction.
5. Clone the analysis task and run nested CV.
6. Retune `c_matrix` on all rows.
7. Fit final raw/stability modes.
8. Plot validation and stability results.
9. Inspect the results in memory.

# Setup

Bootstrap summaries in this workflow are descriptive uncertainty and stability
outputs, not null-hypothesis tests. The built-in permutation options are
conditional diagnostics. A confirmatory permutation analysis must rerun the
complete preprocessing, selection, tuning, and fitting pipeline using the
study's valid exchangeability units. `mb_permutation_test()` provides that
complete-analysis engine; `mbspls_permutation_test()` provides a fixed,
pre-specified omnibus MB-sPLS test. Frozen LC score associations may be tested
with `mb_lc_confirmation_test()` only on genuinely untouched confirmation
observations; its Holm-adjusted p-values concern replication, not generic LC2+
population rank. See `STATISTICAL_VALIDITY.md` and
`MODEL_CARD.md` in the installed package.

```{r setup, message=FALSE}
library(mlr3)
library(mlr3pipelines)
library(mlr3tuning)
library(mlr3cluster)
library(mlr3learners)
library(mlr3mbspls)
library(data.table)
library(ggplot2)

set.seed(42)

# Runtime-focused defaults for an executable package vignette.
# Increase them for an analysis and choose the permutation count from the required
# Monte Carlo precision; valid inference also requires the design rules above.
cfg = list(
  ncomp = 2L,
  centers = 2L,
  inner_folds = 2L,
  outer_folds = 2L,
  tuner_budget = 2L,
  n_perm = 99L,
  n_perm_tuning = 20L,
  bootstrap_B = 20L,
  bootstrap_summary_B = 199L,
  confirmation_n = 80L,
  frequency_threshold = 0.5,
  perf_metric = "mac"
)
```

# 1) Load Packaged Synthetic Task

This task is included in the package and is designed specifically for reproducible MB-sPLS workflows.
Start by checking dimensions and the stored block mapping.

```{r input}
task_source = tsk("mbspls_synthetic_blocks")
dt_demo = as.data.table(task_source$data(cols = task_source$feature_names))
blocks = task_source$block_features()

dim(dt_demo)
head(dt_demo)
blocks
```

# 2) Supported Permutation Inference

The following sections execute every statistically supported test route. They
do not enable the graph's component-wise training permutation or
prediction-side permutation flags: those keep fitted choices fixed and are
conditional diagnostics, not complete-pipeline significance tests in this
workflow.

The 99 permutations used here keep the rendered example brief and give a
minimum attainable p-value of 0.01. Use at least 999 for a routine 0.05 analysis
and more when the required Monte Carlo precision or multiplicity-adjusted
decision threshold demands it.

## Fixed MB-sPLS omnibus and target-block tests

`mbspls_permutation_test()` implements four pre-specified statistics. The two
`global_*` tests address complete block independence. The two `target_*` tests
address independence of the pre-specified target `block_c` from the remaining
blocks. Each call repeats standardisation and the complete sequential MB-sPLS
fit on every shuffle.

```{r fixed-permutation-tests}
raw_blocks = lapply(blocks, function(cols) {
  as.matrix(dt_demo[, ..cols])
})

mb_test_specs = list(
  global_lc1 = list(statistic = "global_lc1", target_block = NULL, seed = 101L),
  global_sum = list(statistic = "global_sum", target_block = NULL, seed = 102L),
  target_lc1 = list(statistic = "target_lc1", target_block = "block_c", seed = 103L),
  target_sum = list(statistic = "target_sum", target_block = "block_c", seed = 104L)
)

mb_tests = lapply(names(mb_test_specs), function(test_name) {
  specification = mb_test_specs[[test_name]]
  arguments = list(
    blocks = raw_blocks,
    ncomp = cfg$ncomp,
    statistic = specification$statistic,
    n_perm = cfg$n_perm,
    seed = specification$seed,
    analysis_seed = 11L
  )
  if (!is.null(specification$target_block)) {
    arguments$target_block = specification$target_block
  }
  do.call(mbspls_permutation_test, arguments)
})
names(mb_tests) = names(mb_test_specs)

invisible(lapply(mb_tests, print))

mb_test_results = rbindlist(lapply(names(mb_tests), function(test_name) {
  test = mb_tests[[test_name]]
  data.table(
    test = test_name,
    observed = test$statistic,
    p_value = test$p_value,
    exceedances = test$exceedances,
    n_perm = test$n_perm
  )
}))
mb_test_results
```

These four calls expose the complete API in one vignette. A real analysis must
pre-specify the statistic that matches its scientific null. Running several
alternatives and reporting the smallest p-value requires an appropriate
multiplicity strategy.

## Complete-analysis test with data-dependent selection

When preprocessing, filtering, tuning, or feature selection depends on the
tested block alignment, it belongs inside `mb_permutation_test()`. This compact
callback refits scaling and reselects the largest absolute cross-block feature
correlation for the observed data and every permutation.

```{r complete-analysis-test}
complete_analysis = function(current_blocks) {
  scaled_blocks = lapply(current_blocks, scale)
  block_pairs = combn(names(scaled_blocks), 2L, simplify = FALSE)

  associations = rbindlist(lapply(block_pairs, function(block_pair) {
    left = scaled_blocks[[block_pair[[1L]]]]
    right = scaled_blocks[[block_pair[[2L]]]]
    feature_pairs = expand.grid(
      left = colnames(left),
      right = colnames(right),
      stringsAsFactors = FALSE
    )
    data.table(
      block_pair = paste(block_pair, collapse = " - "),
      feature_pair = paste(feature_pairs$left, feature_pairs$right, sep = " - "),
      association = as.vector(abs(stats::cor(left, right)))
    )
  }))

  selected = associations[which.max(association)]
  list(
    statistic = selected$association,
    selected_pair = paste(selected$block_pair, selected$feature_pair, sep = ": ")
  )
}

complete_test = mb_permutation_test(
  blocks = raw_blocks,
  analysis = complete_analysis,
  permute_blocks = names(raw_blocks)[-1L],
  n_perm = cfg$n_perm,
  seed = 105L,
  analysis_seed = 11L,
  null_hypothesis = paste(
    "No cross-block feature association is present under independent",
    "row exchangeability."
  )
)

complete_test
complete_test$observed_result$selected_pair
```

## Permutation null-distribution plots

Each vertical line is the observed statistic; each histogram is the sampled
null distribution generated by the corresponding complete refit.

```{r permutation-null-plots, fig.height=7, fig.width=8}
all_permutation_tests = c(mb_tests, list(adaptive_max = complete_test))
permutation_null = rbindlist(lapply(names(all_permutation_tests), function(test_name) {
  data.table(
    test = test_name,
    null_statistic = all_permutation_tests[[test_name]]$null_distribution
  )
}))
permutation_observed = rbindlist(lapply(names(all_permutation_tests), function(test_name) {
  data.table(
    test = test_name,
    observed = all_permutation_tests[[test_name]]$statistic
  )
}))

permutation_plot = ggplot(
  permutation_null,
  aes(x = null_statistic)
) +
  geom_histogram(bins = 18L, fill = "#4472C4", colour = "white") +
  geom_vline(
    data = permutation_observed,
    aes(xintercept = observed),
    colour = "#B22222",
    linewidth = 0.8
  ) +
  facet_wrap(vars(test), scales = "free", ncol = 2L) +
  labs(
    x = "Statistic under the null",
    y = "Permutation count",
    title = "Complete-refit permutation distributions",
    subtitle = "Red line: observed statistic"
  ) +
  theme_minimal(base_size = 11)
permutation_plot
```

## LC tests in independent confirmation data

Individual frozen LC score associations can be tested only in genuinely
untouched confirmation observations. The following simulated confirmation
scores are independent of the earlier fitting by construction: LC1 shares a
signal across blocks, while LC2 is null. In a real study, every score definition
must instead come from a completely frozen discovery transformation.

```{r lc-confirmation-test}
set.seed(20260831L)
confirmation_lc1 = stats::rnorm(cfg$confirmation_n)
confirmation_scores = list(
  block_a = cbind(
    LC_01 = confirmation_lc1 + stats::rnorm(cfg$confirmation_n, sd = 0.08),
    LC_02 = stats::rnorm(cfg$confirmation_n)
  ),
  block_b = cbind(
    LC_01 = confirmation_lc1 + stats::rnorm(cfg$confirmation_n, sd = 0.08),
    LC_02 = stats::rnorm(cfg$confirmation_n)
  ),
  block_c = cbind(
    LC_01 = confirmation_lc1 + stats::rnorm(cfg$confirmation_n, sd = 0.08),
    LC_02 = stats::rnorm(cfg$confirmation_n)
  )
)

lc_confirmation = mb_lc_confirmation_test(
  scores = confirmation_scores,
  independent_confirmation = TRUE,
  permute_blocks = c("block_b", "block_c"),
  n_perm = cfg$n_perm,
  seed = 106L,
  keep_null = TRUE
)
lc_confirmation
```

Raw p-values use `(b + 1) / (B + 1)` and the displayed decisions use Holm
adjustment across the complete two-LC family. These are confirmation claims for
fixed scores, not evidence that the population cross-block rank is at least two.

```{r lc-confirmation-plot, fig.height=4.5, fig.width=7}
lc_null = rbindlist(lapply(seq_along(lc_confirmation$tests), function(component) {
  data.table(
    component = lc_confirmation$results$component[[component]],
    null_statistic = lc_confirmation$tests[[component]]$null_distribution
  )
}))
lc_observed = as.data.table(lc_confirmation$results)[
  , .(component, observed = statistic)
]

lc_confirmation_plot = ggplot(lc_null, aes(x = null_statistic)) +
  geom_histogram(bins = 18L, fill = "#70AD47", colour = "white") +
  geom_vline(
    data = lc_observed,
    aes(xintercept = observed),
    colour = "#B22222",
    linewidth = 0.8
  ) +
  facet_wrap(vars(component), scales = "free") +
  labs(
    x = "Confirmation statistic under the null",
    y = "Permutation count",
    title = "Frozen LC confirmation tests",
    subtitle = "Red line: observed fixed-score association"
  ) +
  theme_minimal(base_size = 11)
lc_confirmation_plot
```

# 3) Bootstrap Uncertainty Is Descriptive

An ordinary paired bootstrap estimates sampling uncertainty around the observed
association. It does not generate a null distribution. The shared summary
therefore reports its estimate, bias, standard error, and interval while keeping
the hypothesis-test p-value explicitly unavailable.

```{r bootstrap-summary}
bootstrap_observed = abs(stats::cor(
  confirmation_scores$block_a[, "LC_01"],
  confirmation_scores$block_b[, "LC_01"]
))

set.seed(20260831L)
bootstrap_replicates = replicate(cfg$bootstrap_summary_B, {
  rows = sample.int(cfg$confirmation_n, replace = TRUE)
  abs(stats::cor(
    confirmation_scores$block_a[rows, "LC_01"],
    confirmation_scores$block_b[rows, "LC_01"]
  ))
})

bootstrap_summary = mb_bootstrap_summary(
  replicates = bootstrap_replicates,
  observed = bootstrap_observed,
  conf = 0.95,
  type = "percentile"
)
bootstrap_summary
```

```{r bootstrap-plot, fig.height=4.5, fig.width=7}
bootstrap_plot_data = data.table(estimate = bootstrap_replicates)
bootstrap_plot = ggplot(bootstrap_plot_data, aes(x = estimate)) +
  geom_histogram(bins = 24L, fill = "#8064A2", colour = "white") +
  geom_vline(
    xintercept = bootstrap_summary$estimate,
    colour = "#B22222",
    linewidth = 0.8
  ) +
  geom_vline(
    xintercept = c(bootstrap_summary$conf_low, bootstrap_summary$conf_high),
    linetype = "dashed",
    linewidth = 0.6
  ) +
  labs(
    x = "Bootstrap LC1 association",
    y = "Bootstrap count",
    title = "Descriptive bootstrap uncertainty",
    subtitle = "Red: observed estimate; dashed: percentile interval"
  ) +
  theme_minimal(base_size = 11)
bootstrap_plot
```

# 4) Inspect Task Quality

Before fitting any model, run a quick QC summary once. This is especially helpful in multi-omics, clinical, neuroimaging, and socioeconomic workflows where block imbalance or missingness may drive downstream results.

```{r qc}
qc = task_source$overview()
qc$overview
qc$blocks
qc$issues
```

# 5) Define Site Correction

The block assignment already lives on the task, so only the site-correction policy needs to be declared here.

```{r blocks}
site_correction = list(block_a = "site_batch", block_b = "site_batch", block_c = "site_batch")
site_correction_methods = list(block_a = "partial_corr", block_b = "partial_corr", block_c = "partial_corr")

lengths(blocks)
```

# 6) Clone the Clustering Task

The packaged synthetic object is already a `TaskMultiBlock`, so for analysis you can usually just clone it.

```{r task}
task_train = task_source$clone(deep = TRUE)

task_train
length(task_train$feature_names)
```

# 7) Nested CV (Primary Validation)

This stage estimates held-out performance; it does not by itself perform a
null-hypothesis test.
Bootstrap selection is intentionally disabled here to evaluate core model generalization under nested resampling.

```{r nested-cv}
gl_nested = ppl(
  "mbspls_graph_learner",
  learner = lrn("clust.kmeans", centers = cfg$centers),
  task = task_train,
  site_correction = site_correction,
  site_correction_methods = site_correction_methods,
  ncomp = cfg$ncomp,
  performance_metric = cfg$perf_metric,
  permutation_test = FALSE,
  bootstrap = FALSE,
  bootstrap_selection = FALSE,
  B = 1L,
  val_test = "none"
)

rs_outer = rsmp("cv", folds = cfg$outer_folds)
rs_inner = rsmp("cv", folds = cfg$inner_folds)
rs_outer$instantiate(task_train)

nested_cv = mbspls_nested_cv(
  task = task_train,
  graphlearner = gl_nested,
  rs_outer = rs_outer,
  rs_inner = rs_inner,
  ncomp = cfg$ncomp,
  tuner_budget = cfg$tuner_budget,
  tuning_early_stop = FALSE,
  performance_metric = cfg$perf_metric,
  val_test = "none",
  n_perm_tuning = cfg$n_perm_tuning,
  store_payload = TRUE
)

nested_cv$summary_table
```

The nested summary is the key validation output and should be inspected before any final-model interpretation.

# 8) Retune c_matrix on All Rows

After nested validation, retune on the full training task to obtain a single final sparsity pattern (`c_matrix_final`).
This matrix is then reused across all final modes for comparability.

```{r tune}
gl_tune = ppl(
  "mbspls_graph_learner",
  learner = lrn("clust.kmeans", centers = cfg$centers),
  task = task_train,
  site_correction = site_correction,
  site_correction_methods = site_correction_methods,
  ncomp = cfg$ncomp,
  performance_metric = cfg$perf_metric,
  permutation_test = FALSE,
  bootstrap = FALSE,
  bootstrap_selection = FALSE,
  B = 1L,
  val_test = "none"
)

tuner = TunerSeqMBsPLS$new(
  tuner = "random_search",
  budget = cfg$tuner_budget,
  resampling = rsmp("cv", folds = cfg$inner_folds),
  parallel = "none",
  early_stopping = FALSE,
  n_perm = cfg$n_perm_tuning,
  performance_metric = cfg$perf_metric
)

instance = ti(
  task = task_train,
  learner = gl_tune,
  resampling = rsmp("insample"),
  measure = msr("mbspls.mac_evwt"),
  terminator = trm("evals", n_evals = 1)
)

tuner$optimize(instance)
c_matrix_final = instance$result$learner_param_vals[[1]]$c_matrix
c_matrix_final
```

At this point, `c_matrix_final` is the most important object to inspect.
It controls feature sparsity per block and per component in the final fits.

For supervised analyses, the package also provides `mbspls_model_summary()` to extract compact reporting tables from fitted MB-sPLS, MB-sPLS-XY, or MB-sPCA models.

# 9) Final Fits (raw / stable_ci / stable_frequency)

Final fitting creates three representations:

- `raw`: direct MB-sPLS weights
- `stable_ci`: CI-filtered bootstrap-stable weights
- `stable_frequency`: frequency-filtered bootstrap-stable weights

Bootstrap selection is activated here for stability reporting. The preceding
outer-fold estimates evaluate the raw representation, not these additional
selection procedures. To estimate performance for a stability-filtered
pipeline, repeat bootstrap selection inside every relevant training split and
evaluate that complete pipeline on held-out observations.

```{r final-fits}
modes = list(
  raw = list(selection = "none", predict_weights = "raw", threshold = NA_real_),
  stable_ci = list(selection = "ci", predict_weights = "stable_ci", threshold = NA_real_),
  stable_frequency = list(selection = "frequency", predict_weights = "stable_frequency", threshold = 0.5)
)

final_models = list()
final_train_summary = list()
final_train_predictions = list()

for (mode_name in names(modes)) {
  mode_spec = modes[[mode_name]]

  gl_final = ppl(
    "mbspls_graph_learner",
    learner = lrn("clust.kmeans", centers = cfg$centers),
    blocks = blocks,
    site_correction = site_correction,
    site_correction_methods = site_correction_methods,
    ncomp = cfg$ncomp,
    performance_metric = cfg$perf_metric,
    permutation_test = FALSE,
    bootstrap = TRUE,
    bootstrap_selection = mode_spec$selection != "none",
    B = cfg$bootstrap_B,
    val_test = "none",
    selection_method = if (mode_spec$selection == "frequency") "frequency" else "ci",
    frequency_threshold = if (is.na(mode_spec$threshold)) cfg$frequency_threshold else mode_spec$threshold
  )

  gl_final$param_set$values$mbspls.c_matrix = c_matrix_final
  if (!is.null(gl_final$graph$pipeops$mbspls)) {
    gl_final$graph$pipeops$mbspls$param_set$values$c_matrix = c_matrix_final
    gl_final$graph$pipeops$mbspls$param_set$values$predict_weights = mode_spec$predict_weights
    gl_final$graph$pipeops$mbspls$param_set$values$store_train_blocks = TRUE
  }

  gl_final$train(task_train)
  pred_train = gl_final$predict(task_train)

  state = gl_final$model$mbspls
  mac_comp = as.numeric(state$obj_vec)
  ev_comp = as.numeric(state$ev_comp)

  summary_dt = data.table(
    mode = mode_name,
    component = sprintf("LC_%02d", seq_along(mac_comp)),
    train_objective = mac_comp,
    train_ev = ev_comp
  )

  final_models[[mode_name]] = gl_final
  final_train_summary[[mode_name]] = summary_dt
  final_train_predictions[[mode_name]] = data.table(
    row_id = pred_train$row_ids,
    cluster = as.character(pred_train$partition)
  )
}

final_train_summary = rbindlist(final_train_summary)
final_train_summary
```

Comparing the three mode summaries is a practical way to assess how stability filtering changes component strength and downstream clustering.

# 10) Visualize Validation and Stability

The outer-fold plot shows assessment performance only. It is not a permutation
test and is not interchangeable with the null-distribution plots above.

```{r nested-cv-plot, fig.height=4.5, fig.width=7}
nested_plot_data = nested_cv$results[
  , .(split = factor(split), assessment_score = measure_test)
]
nested_plot = ggplot(
  nested_plot_data,
  aes(x = split, y = assessment_score)
) +
  geom_col(fill = "#5B9BD5", width = 0.7) +
  geom_hline(
    yintercept = mean(nested_plot_data$assessment_score),
    linetype = "dashed",
    linewidth = 0.6
  ) +
  labs(
    x = "Outer fold",
    y = "EV-weighted MAC",
    title = "Nested-CV assessment performance",
    subtitle = "Dashed line: mean across outer folds"
  ) +
  theme_minimal(base_size = 11)
nested_plot
```

The next plot shows sign-aligned bootstrap means, percentile intervals, and
selection-frequency opacity. These are stability summaries rather than
feature-level significance tests.

```{r weight-stability-plot, fig.height=7.5, fig.width=12}
weight_stability_plot = mbspls_plot_block_weight_ci(
  final_models$stable_ci,
  source = "bootstrap",
  ci_filter = "none",
  top_n = 4L,
  add_block_rule = FALSE,
  alpha_by_stability = TRUE
)
weight_stability_plot
```

Explained variance provides a complementary component-level description of the
same final model.

```{r variance-plot, fig.height=4.5, fig.width=7}
variance_plot = autoplot(
  final_models$stable_ci,
  type = "mbspls_variance",
  source = "bootstrap",
  show_total = TRUE
)
variance_plot
```

# 11) Inspect Results in Memory

The analysis does not write results to the working directory. Rendering uses
temporary plot files; nested-CV results, fitted models, summaries, predictions,
and runtime versions remain available in the current R session.

```{r inspect-results}
names(final_models)
head(final_train_summary)
head(final_train_predictions$raw)

runtime_versions = data.table(
  package = c("R", "mlr3", "mlr3mbspls"),
  version = c(
    as.character(getRversion()),
    as.character(packageVersion("mlr3")),
    as.character(packageVersion("mlr3mbspls"))
  )
)
runtime_versions
```

# 12) Objects to Inspect During Development

At any point, inspect objects such as:

- `dt_demo`
- `blocks`
- `mb_test_results`
- `complete_test`
- `lc_confirmation$results`
- `bootstrap_summary`
- `task_train`
- `nested_cv$summary_table`
- `c_matrix_final`
- `final_train_summary`
- `final_train_predictions`
- `final_models$raw$model$mbspls`

This keeps the workflow easy to debug while remaining fully executable end-to-end.
