---
title: "Quick Start with mlr3mbspls"
author: "Stefan Coors, Clara Vetter"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start with mlr3mbspls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Goal

This vignette shows a straightforward, script-style MB-sPLS workflow:

- Uses the packaged synthetic `TaskMultiBlock` `mbspls_synthetic_blocks`.
- Notes the matching supervised toy tasks `mbspls_synthetic_classif` and `mbspls_synthetic_regr`.
- Generic naming only.
- No helper function definitions.
- Sequential objects that are easy to inspect in an interactive R session.

The sequence mirrors a typical analysis run:

1. Load the packaged synthetic multi-block task.
2. Reuse its block mapping and define site correction.
3. Clone the analysis task.
4. Run nested CV.
5. Retune `c_matrix` on all rows.
6. Fit final raw/stability modes.
7. Save outputs.

# Setup
# Goal

This vignette shows a straightforward, script-style MB-sPLS workflow:

- Uses the packaged synthetic `TaskMultiBlock` `mbspls_synthetic_blocks`.
- Notes the matching supervised toy tasks `mbspls_synthetic_classif` and `mbspls_synthetic_regr`.
- Generic naming only.
- No helper function definitions.
- Sequential objects that are easy to inspect in an interactive R session.

The sequence mirrors a typical analysis run:

1. Load the packaged synthetic multi-block task.
2. Reuse its block mapping and define site correction.
3. Clone the analysis task.
4. Run nested CV.
5. Retune `c_matrix` on all rows.
6. Fit final raw/stability modes.
7. Save outputs.

# Setup

```{r setup, message=FALSE}
library(mlr3)
library(mlr3pipelines)
library(mlr3tuning)
library(mlr3cluster)
library(mlr3learners)
library(mlr3mbspls)
library(data.table)
library(ggplot2)

set.seed(42)

# Runtime-focused defaults for a quick, reliable walkthrough.
# Increase these values for deeper inferential analyses.
cfg = list(
  ncomp = 3L,
  centers = 2L,
  inner_folds = 3L,
  outer_folds = 3L,
  tuner_budget = 40L,
  n_perm = 40L,
  n_perm_tuning = 40L,
  val_test_n = 40L,
  bootstrap_B = 40L,
  frequency_threshold = 0.5,
  perf_metric = "mac"
)
```

# 1) Load Packaged Synthetic Task

This task is included in the package and is designed specifically for reproducible MB-sPLS workflows.
Start by checking dimensions and the stored block mapping.

```{r input}
task_source = tsk("mbspls_synthetic_blocks")
dt_demo = as.data.table(task_source$data(cols = task_source$feature_names))
blocks = task_source$block_features()

dim(dt_demo)
head(dt_demo)
blocks
```

# 2) Inspect Task Quality

Before fitting any model, run a quick QC summary once. This is especially helpful in multi-omics, clinical, neuroimaging, and socioeconomic workflows where block imbalance or missingness may drive downstream results.

```{r qc}
qc = mb_task_overview(task_source)
qc$overview
qc$blocks
qc$issues
```

# 3) Define Site Correction

The block assignment already lives on the task, so only the site-correction policy needs to be declared here.

```{r blocks}
site_correction = list(block_a = "site_batch", block_b = "site_batch", block_c = "site_batch")
site_correction_methods = list(block_a = "partial_corr", block_b = "partial_corr", block_c = "partial_corr")

lengths(blocks)
```

# 4) Clone the Clustering Task

The packaged synthetic object is already a `TaskMultiBlock`, so for analysis you can usually just clone it.

```{r task}
task_train = TaskMultiBlock(
  task_source,
  id = "mbspls_synthetic_blocks_analysis"
)

task_train
length(task_train$feature_names)
```

# 5) Nested CV (Primary Validation)

This stage performs the primary inferential validation.
Bootstrap selection is intentionally disabled here to evaluate core model generalization under nested resampling.

```{r nested-cv}
gl_nested = ppl(
  "mbspls_graph_learner",
  learner = lrn("clust.kmeans", centers = cfg$centers),
  task = task_train,
  site_correction = site_correction,
  site_correction_methods = site_correction_methods,
  ncomp = cfg$ncomp,
  performance_metric = cfg$perf_metric,
  permutation_test = TRUE,
  n_perm = cfg$n_perm,
  bootstrap = FALSE,
  bootstrap_selection = FALSE,
  B = 1L,
  val_test = "permutation",
  val_test_n = cfg$val_test_n
)

rs_outer = rsmp("cv", folds = cfg$outer_folds)
rs_inner = rsmp("cv", folds = cfg$inner_folds)
rs_outer$instantiate(task_train)

nested_cv = mbspls_nested_cv(
  task = task_train,
  graphlearner = gl_nested,
  rs_outer = rs_outer,
  rs_inner = rs_inner,
  ncomp = cfg$ncomp,
  tuner_budget = cfg$tuner_budget,
  tuning_early_stop = TRUE,
  performance_metric = cfg$perf_metric,
  val_test = "permutation",
  val_test_n = cfg$val_test_n,
  n_perm_tuning = cfg$n_perm_tuning,
  store_payload = TRUE
)

nested_cv$summary_table
```

The nested summary is the key validation output and should be inspected before any final-model interpretation.

# 6) Retune c_matrix on All Rows

After nested validation, retune on the full training task to obtain a single final sparsity pattern (`c_matrix_final`).
This matrix is then reused across all final modes for comparability.

```{r tune}
gl_tune = ppl(
  "mbspls_graph_learner",
  learner = lrn("clust.kmeans", centers = cfg$centers),
  task = task_train,
  site_correction = site_correction,
  site_correction_methods = site_correction_methods,
  ncomp = cfg$ncomp,
  performance_metric = cfg$perf_metric,
  permutation_test = TRUE,
  n_perm = cfg$n_perm,
  bootstrap = FALSE,
  bootstrap_selection = FALSE,
  B = 1L,
  val_test = "none"
)

tuner = TunerSeqMBsPLS$new(
  tuner = "random_search",
  budget = cfg$tuner_budget,
  resampling = rsmp("cv", folds = cfg$inner_folds),
  parallel = "none",
  early_stopping = TRUE,
  n_perm = cfg$n_perm_tuning,
  performance_metric = cfg$perf_metric
)

instance = ti(
  task = task_train,
  learner = gl_tune,
  resampling = rsmp("insample"),
  measure = msr("mbspls.mac_evwt"),
  terminator = trm("evals", n_evals = 1)
)

tuner$optimize(instance)
c_matrix_final = instance$result$learner_param_vals[[1]]$c_matrix
c_matrix_final
```

At this point, `c_matrix_final` is the most important object to inspect.
It controls feature sparsity per block and per component in the final fits.

For supervised analyses, the package also provides `mbspls_model_summary()` to extract compact reporting tables from fitted MB-sPLS, MB-sPLS-XY, or MB-sPCA models.

# 7) Final Fits (raw / stable_ci / stable_frequency)

Final fitting creates three representations:

- `raw`: direct MB-sPLS weights
- `stable_ci`: CI-filtered bootstrap-stable weights
- `stable_frequency`: frequency-filtered bootstrap-stable weights

Bootstrap selection is only activated here, where stability-oriented reporting is expected.

```{r final-fits}
output_dir = file.path(
  "analysis_results",
  paste0(format(Sys.time(), "%Y%m%d_%H%M%S"), "_MBSPLS_SYNTHETIC_BLOCKS")
)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

modes = list(
  raw = list(selection = "none", predict_weights = "raw", threshold = NA_real_),
  stable_ci = list(selection = "ci", predict_weights = "stable_ci", threshold = NA_real_),
  stable_frequency = list(selection = "frequency", predict_weights = "stable_frequency", threshold = 0.5)
)

final_models = list()
final_train_summary = list()

for (mode_name in names(modes)) {
  mode_spec = modes[[mode_name]]
  mode_dir = file.path(output_dir, mode_name)
  dir.create(mode_dir, recursive = TRUE, showWarnings = FALSE)

  gl_final = ppl(
    "mbspls_graph_learner",
    learner = lrn("clust.kmeans", centers = cfg$centers),
    blocks = blocks,
    site_correction = site_correction,
    site_correction_methods = site_correction_methods,
    ncomp = cfg$ncomp,
    performance_metric = cfg$perf_metric,
    permutation_test = TRUE,
    n_perm = cfg$n_perm,
    bootstrap = TRUE,
    bootstrap_selection = mode_spec$selection != "none",
    B = cfg$bootstrap_B,
    val_test = "none",
    selection_method = if (mode_spec$selection == "frequency") "frequency" else "ci",
    frequency_threshold = if (is.na(mode_spec$threshold)) cfg$frequency_threshold else mode_spec$threshold
  )

  gl_final$param_set$values$mbspls.c_matrix = c_matrix_final
  if (!is.null(gl_final$graph$pipeops$mbspls)) {
    gl_final$graph$pipeops$mbspls$param_set$values$c_matrix = c_matrix_final
    gl_final$graph$pipeops$mbspls$param_set$values$predict_weights = mode_spec$predict_weights
    gl_final$graph$pipeops$mbspls$param_set$values$store_train_blocks = TRUE
  }

  gl_final$train(task_train)
  pred_train = gl_final$predict(task_train)

  state = gl_final$model$mbspls
  mac_comp = as.numeric(state$obj_vec)
  ev_comp = as.numeric(state$ev_comp)

  summary_dt = data.table(
    mode = mode_name,
    component = sprintf("LC_%02d", seq_along(mac_comp)),
    train_objective = mac_comp,
    train_ev = ev_comp
  )

  fwrite(summary_dt, file.path(mode_dir, "train_summary.csv"))
  fwrite(data.table(row_id = pred_train$row_ids, cluster = as.character(pred_train$partition)),
    file.path(mode_dir, "clusters_train.csv"))

  saveRDS(gl_final, file.path(mode_dir, "graphlearner.rds"))
  saveRDS(state, file.path(mode_dir, "train_state.rds"))

  final_models[[mode_name]] = gl_final
  final_train_summary[[mode_name]] = summary_dt
}

rbindlist(final_train_summary)
```

Comparing the three mode summaries is a practical way to assess how stability filtering changes component strength and downstream clustering.

# 7) Save Nested CV and Session Info

Write compact tabular outputs plus serialized objects for reproducibility and downstream reporting.
Saving `session_info.txt` helps track package versions and runtime context.

```{r save-outputs}
nested_dir = file.path(output_dir, "nested_cv")
dir.create(nested_dir, recursive = TRUE, showWarnings = FALSE)

fwrite(as.data.table(nested_cv$summary_table), file.path(nested_dir, "nested_cv_summary.csv"))

nested_results_csv = copy(as.data.table(nested_cv$results))
for (col in names(nested_results_csv)[vapply(nested_results_csv, is.list, logical(1))]) {
  nested_results_csv[, (col) := vapply(get(col), function(x) paste(as.character(x), collapse = ";"), character(1))]
}
fwrite(nested_results_csv, file.path(nested_dir, "nested_cv_results.csv"))

saveRDS(nested_cv, file.path(nested_dir, "nested_cv_object.rds"))
writeLines(capture.output(sessionInfo()), con = file.path(output_dir, "session_info.txt"))

output_dir
```

# 8) Objects to Inspect During Development

At any point, inspect objects such as:

- `dt_demo`
- `blocks`
- `task_train`
- `nested_cv$summary_table`
- `c_matrix_final`
- `final_models$raw$model$mbspls$state`

This keeps the workflow easy to debug while remaining fully executable end-to-end.
