---
title: "Quick Start with mlr3mbspls"
author: "Stefan Coors, Clara Vetter"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start with mlr3mbspls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  dev = "svg",
  eval = TRUE,
  fig.align = "center",
  fig.height = 4.5,
  fig.path = file.path(tempdir(), "mlr3mbspls-quickstart", ""),
  fig.width = 7
)
```

# Goal

This vignette shows a straightforward, script-style MB-sPLS workflow:

- Uses the packaged synthetic `TaskMultiBlock` `mbspls_synthetic_blocks`.
- Notes the matching supervised toy tasks `mbspls_synthetic_classif` and `mbspls_synthetic_regr`.
- Sequential objects that are easy to inspect in an interactive R session.
- Executes every supported permutation-inference route and shows its output.

The sequence mirrors a typical analysis run:

1. Load the packaged synthetic multi-block task.
2. Run and plot the supported permutation tests.
3. Show descriptive bootstrap uncertainty.
4. Inspect task quality.
5. Define site correction.
6. Clone the analysis task.
7. Define raw and stability-selection pipelines and validate each with nested
   CV.
8. Retune `c_matrix` on all rows.
9. Fit the final pipelines and compare what the selection modes change.
10. Plot validation and stability results.
11. Inspect the results in memory.

# Setup

Bootstrap summaries in this workflow are descriptive uncertainty and stability
outputs, not null-hypothesis tests. The built-in permutation options are
conditional diagnostics. A confirmatory permutation analysis must rerun the
complete preprocessing, selection, tuning, and fitting pipeline using the
study's valid exchangeability units. `mb_permutation_test()` provides that
complete-analysis engine; `mbspls_permutation_test()` provides a fixed,
pre-specified omnibus MB-sPLS test. Frozen LC score associations may be tested
with `mb_lc_confirmation_test()` only on genuinely untouched confirmation
observations; with `reference_signs` from discovery its Holm-adjusted p-values
concern directional replication, not generic LC2+ population rank. See
`STATISTICAL_VALIDITY.md` and `MODEL_CARD.md` in the installed package.

```{r setup, message=FALSE}
library(mlr3)
library(mlr3pipelines)
library(mlr3tuning)
library(mlr3cluster)
library(mlr3learners)
library(mlr3mbspls)
library(data.table)
library(ggplot2)

lgr::lgr$set_threshold("warn")
lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("bbotk")$set_threshold("warn")

set.seed(42)

# Runtime-focused defaults for an executable package vignette.
# Increase them for an analysis and choose the permutation count from the required
# Monte Carlo precision; valid inference also requires the design rules above.
cfg = list(
  ncomp = 2L,
  centers = 2L,
  inner_folds = 2L,
  outer_folds = 2L,
  tuner_budget = 2L,
  n_perm = 99L,
  bootstrap_B = 20L,
  bootstrap_summary_B = 199L,
  confirmation_n = 80L,
  frequency_threshold = 0.5,
  perf_metric = "mac",
  seed_bootstrap = 12L,
  seed_tuning = 7L,
  seed_final = 8L
)
```

# 1) Load Packaged Synthetic Task

This task is included in the package and is designed specifically for reproducible MB-sPLS workflows.
Start by checking dimensions, the stored block mapping, and the site variable.

```{r input}
task_source = tsk("mbspls_synthetic_blocks")
dt_demo = as.data.table(task_source$data(cols = task_source$feature_names))
blocks = task_source$block_features()
site = task_source$data(cols = "site_batch")[[1L]]

dim(dt_demo)
head(dt_demo)
blocks
table(site)
```

# 2) Supported Permutation Inference

The following sections execute every statistically supported test route. They
do not enable the graph's component-wise training permutation or
prediction-side permutation flags: those keep fitted choices fixed and are
conditional diagnostics, not complete-pipeline significance tests in this
workflow.

The 99 permutations used here keep the rendered example brief and give a
minimum attainable p-value of 0.01. Use at least 999 for a routine 0.05 analysis
and more when the required Monte Carlo precision or multiplicity-adjusted
decision threshold demands it. With few exchangeable units or small strata, the
number of distinct permutations, not `n_perm`, limits the attainable p-value to
about one over that number. Every result stores this permutation group size
(`log_permutation_group_size`), and a warning is raised when `n_perm` reaches
it.

## Site as a nuisance variable

The synthetic generator adds the same site shift to every feature of every
block, so `site_batch` alone creates dependence between the blocks, and later
sections correct all blocks for it. An unrestricted shuffle tests marginal
block independence, which a shared site or scanner shift can reject on its own.
The tests below therefore permute rows within sites (`strata = site`). They
test independence of the blocks conditional on site and assume that rows are
exchangeable within each site. Residualising the blocks on site and then
shuffling freely is not automatically valid; see the permutation-test section
of `STATISTICAL_VALIDITY.md` and the details of `mb_permutation_test()`.

## Fixed MB-sPLS omnibus and target-block tests

`mbspls_permutation_test()` implements four pre-specified statistics. The two
`global_*` tests address block independence. The two `target_*` tests address
independence of the pre-specified target `block_c` from the remaining blocks.
Standardisation is refitted inside every observed and permuted fit. The `*_sum`
statistics refit every component on every shuffle; the `*_lc1` statistics
refit LC1 only, because sequential extraction fits LC1 before any deflation,
and report later components from one fit to the observed data.

```{r fixed-permutation-tests}
raw_blocks = lapply(blocks, function(cols) {
  as.matrix(dt_demo[, ..cols])
})

mb_test_specs = list(
  global_lc1 = list(statistic = "global_lc1", target_block = NULL, seed = 101L),
  global_sum = list(statistic = "global_sum", target_block = NULL, seed = 102L),
  target_lc1 = list(statistic = "target_lc1", target_block = "block_c", seed = 103L),
  target_sum = list(statistic = "target_sum", target_block = "block_c", seed = 104L)
)

mb_tests = lapply(names(mb_test_specs), function(test_name) {
  specification = mb_test_specs[[test_name]]
  arguments = list(
    blocks = raw_blocks,
    ncomp = cfg$ncomp,
    statistic = specification$statistic,
    n_perm = cfg$n_perm,
    strata = site,
    seed = specification$seed
  )
  if (!is.null(specification$target_block)) {
    arguments$target_block = specification$target_block
  }
  do.call(mbspls_permutation_test, arguments)
})
names(mb_tests) = names(mb_test_specs)

invisible(lapply(mb_tests, print))

mb_test_results = rbindlist(lapply(names(mb_tests), function(test_name) {
  test = mb_tests[[test_name]]
  data.table(
    test = test_name,
    observed = test$statistic,
    p_value = test$p_value,
    exceedances = test$exceedances,
    n_perm = test$n_perm,
    mc_conf_high = test$monte_carlo_conf_high,
    n_strata = test$n_strata
  )
}))
mb_test_results
```

The printed header labels these results as fixed-specification tests: their
p-values are valid because the statistic, `ncomp`, the default dense sparsity
budgets, standardisation, and solver settings were fixed without looking at
the tested alignment. Sparsity or component counts tuned on the same data
require the complete-analysis route below.

The MB-sPLS solver starts from a deterministic cross-covariance
initialisation, so `analysis_seed` does not change these statistics. The calls
keep its default, and the printed seeds record both the permutation seed and
the analysis seed for the report. When an analysis callback contains
stochastic steps, the analysis seed is part of the definition of the
statistic: fix it before looking at the data and never choose among seeds
after seeing results. The analysis seed selects the L'Ecuyer-CMRG generator,
so callbacks that use forked parallelism remain reproducible.

The Monte Carlo line gives the exact 95% Clopper-Pearson interval for the
exceedance probability `b / B`. Report this interval as the Monte Carlo
precision: with no exceedance in 99 permutations its upper bound is still about
0.037.

These four calls expose the complete API in one vignette. A real analysis must
pre-specify the statistic that matches its scientific null. Running several
alternatives and reporting the smallest p-value requires an appropriate
multiplicity strategy.

## Complete-analysis test with data-dependent selection

When preprocessing, filtering, tuning, or feature selection depends on the
tested block alignment, it belongs inside `mb_permutation_test()`. This compact
callback refits scaling and reselects the largest absolute cross-block feature
correlation for the observed data and every permutation. It draws no random
numbers, so the analysis seed is irrelevant here.

```{r complete-analysis-test}
complete_analysis = function(current_blocks) {
  scaled_blocks = lapply(current_blocks, scale)
  block_pairs = combn(names(scaled_blocks), 2L, simplify = FALSE)

  associations = rbindlist(lapply(block_pairs, function(block_pair) {
    left = scaled_blocks[[block_pair[[1L]]]]
    right = scaled_blocks[[block_pair[[2L]]]]
    feature_pairs = expand.grid(
      left = colnames(left),
      right = colnames(right),
      stringsAsFactors = FALSE
    )
    data.table(
      block_pair = paste(block_pair, collapse = " - "),
      feature_pair = paste(feature_pairs$left, feature_pairs$right, sep = " - "),
      association = as.vector(abs(stats::cor(left, right)))
    )
  }))

  selected = associations[which.max(association)]
  list(
    statistic = selected$association,
    selected_pair = paste(selected$block_pair, selected$feature_pair, sep = ": ")
  )
}

complete_test = mb_permutation_test(
  blocks = raw_blocks,
  analysis = complete_analysis,
  permute_blocks = names(raw_blocks)[-1L],
  n_perm = cfg$n_perm,
  strata = site,
  seed = 105L,
  null_hypothesis = paste(
    "No cross-block feature association is present under row",
    "exchangeability within site."
  )
)

complete_test
complete_test$observed_result$selected_pair
```

## Permutation null-distribution plots

Each vertical line is the observed statistic; each histogram is the sampled
null distribution generated by refitting the corresponding analysis on every
within-site shuffle.

```{r permutation-null-plots, fig.height=7, fig.width=8}
all_permutation_tests = c(mb_tests, list(adaptive_max = complete_test))
permutation_null = rbindlist(lapply(names(all_permutation_tests), function(test_name) {
  data.table(
    test = test_name,
    null_statistic = all_permutation_tests[[test_name]]$null_distribution
  )
}))
permutation_observed = rbindlist(lapply(names(all_permutation_tests), function(test_name) {
  data.table(
    test = test_name,
    observed = all_permutation_tests[[test_name]]$statistic
  )
}))

permutation_plot = ggplot(
  permutation_null,
  aes(x = null_statistic)
) +
  geom_histogram(bins = 18L, fill = "#4472C4", colour = "white") +
  geom_vline(
    data = permutation_observed,
    aes(xintercept = observed),
    colour = "#B22222",
    linewidth = 0.8
  ) +
  facet_wrap(vars(test), scales = "free", ncol = 2L) +
  labs(
    x = "Statistic under the null",
    y = "Permutation count",
    title = "Within-site permutation distributions",
    subtitle = "Red line: observed statistic"
  ) +
  theme_minimal(base_size = 11)
permutation_plot
```

## LC tests in independent confirmation data

Individual frozen LC score associations can be tested only in genuinely
untouched confirmation observations. The following simulated confirmation
scores are independent of the earlier fitting by construction: LC1 shares a
signal across blocks, while LC2 is null. They contain no site effect, so rows
are permuted without strata; confirmation scores from a multisite cohort would
need the same design restrictions as above. In a real study, every score
definition must instead come from a completely frozen discovery
transformation.

Frozen weights fix the sign of every block score, so a confirmation
correlation of the opposite sign contradicts discovery rather than replicating
it. `reference_signs` gives the expected sign of each pairwise score
correlation, recycled across LCs here. In a real study, take the signs from the
discovery data, for example from the pairwise correlations of the training
block scores of each frozen LC. Each LC is then tested one-sided in the
discovery direction, and it counts as replicated (`replicated`) only if it is
significant and its observed sign-oriented statistic is positive: under
stratified or whole-unit permutations a pooled correlation of the opposite
sign can still be significant relative to the design. Without
`reference_signs`, the statistic is unsigned and the test establishes
dependence in either direction, not replication.

```{r lc-confirmation-test}
set.seed(20260831L)
confirmation_lc1 = stats::rnorm(cfg$confirmation_n)
confirmation_scores = list(
  block_a = cbind(
    LC_01 = confirmation_lc1 + stats::rnorm(cfg$confirmation_n, sd = 0.08),
    LC_02 = stats::rnorm(cfg$confirmation_n)
  ),
  block_b = cbind(
    LC_01 = confirmation_lc1 + stats::rnorm(cfg$confirmation_n, sd = 0.08),
    LC_02 = stats::rnorm(cfg$confirmation_n)
  ),
  block_c = cbind(
    LC_01 = confirmation_lc1 + stats::rnorm(cfg$confirmation_n, sd = 0.08),
    LC_02 = stats::rnorm(cfg$confirmation_n)
  )
)

lc_confirmation = mb_lc_confirmation_test(
  scores = confirmation_scores,
  independent_confirmation = TRUE,
  permute_blocks = c("block_b", "block_c"),
  reference_signs = c(
    "block_a:block_b" = 1,
    "block_a:block_c" = 1,
    "block_b:block_c" = 1
  ),
  n_perm = cfg$n_perm,
  seed = 106L,
  keep_null = TRUE
)
lc_confirmation
```

Raw p-values use `(b + 1) / (B + 1)` and the displayed decisions use Holm
adjustment across the complete two-LC family. The statistic is the mean
sign-oriented correlation; LC2, which is null by construction, gets a large
p-value here. These are directional confirmation claims for fixed
scores, not evidence that the population cross-block rank is at least two. The
`monte_carlo_conf_low` and `monte_carlo_conf_high` columns give the exact 95%
Clopper-Pearson interval for each exceedance probability, the Monte Carlo
precision to report.

```{r lc-confirmation-plot, fig.height=4.5, fig.width=7}
lc_null = rbindlist(lapply(seq_along(lc_confirmation$tests), function(component) {
  data.table(
    component = lc_confirmation$results$component[[component]],
    null_statistic = lc_confirmation$tests[[component]]$null_distribution
  )
}))
lc_observed = as.data.table(lc_confirmation$results)[
  , .(component, observed = statistic)
]

lc_confirmation_plot = ggplot(lc_null, aes(x = null_statistic)) +
  geom_histogram(bins = 18L, fill = "#70AD47", colour = "white") +
  geom_vline(
    data = lc_observed,
    aes(xintercept = observed),
    colour = "#B22222",
    linewidth = 0.8
  ) +
  facet_wrap(vars(component), scales = "free") +
  labs(
    x = "Mean sign-oriented correlation under the null",
    y = "Permutation count",
    title = "Frozen LC confirmation tests",
    subtitle = "Red line: observed fixed-score association"
  ) +
  theme_minimal(base_size = 11)
lc_confirmation_plot
```

# 3) Bootstrap Uncertainty Is Descriptive

An ordinary paired bootstrap estimates sampling uncertainty around the observed
association. It does not generate a null distribution. The shared summary
therefore reports its estimate, bias, standard error, and interval while keeping
the hypothesis-test p-value explicitly unavailable.

```{r bootstrap-summary}
bootstrap_observed = abs(stats::cor(
  confirmation_scores$block_a[, "LC_01"],
  confirmation_scores$block_b[, "LC_01"]
))

set.seed(20260831L)
bootstrap_replicates = replicate(cfg$bootstrap_summary_B, {
  rows = sample.int(cfg$confirmation_n, replace = TRUE)
  abs(stats::cor(
    confirmation_scores$block_a[rows, "LC_01"],
    confirmation_scores$block_b[rows, "LC_01"]
  ))
})

bootstrap_summary = mb_bootstrap_summary(
  replicates = bootstrap_replicates,
  observed = bootstrap_observed,
  conf = 0.95,
  type = "percentile"
)
bootstrap_summary[c(
  "estimate", "bias", "standard_error", "conf_low", "conf_high",
  "interval_type", "p_value", "p_value_note"
)]
```

```{r bootstrap-plot, fig.height=4.5, fig.width=7}
bootstrap_plot_data = data.table(estimate = bootstrap_replicates)
bootstrap_plot = ggplot(bootstrap_plot_data, aes(x = estimate)) +
  geom_histogram(bins = 24L, fill = "#8064A2", colour = "white") +
  geom_vline(
    xintercept = bootstrap_summary$estimate,
    colour = "#B22222",
    linewidth = 0.8
  ) +
  geom_vline(
    xintercept = c(bootstrap_summary$conf_low, bootstrap_summary$conf_high),
    linetype = "dashed",
    linewidth = 0.6
  ) +
  labs(
    x = "Bootstrap LC1 association",
    y = "Bootstrap count",
    title = "Descriptive bootstrap uncertainty",
    subtitle = "Red: observed estimate; dashed: percentile interval"
  ) +
  theme_minimal(base_size = 11)
bootstrap_plot
```

# 4) Inspect Task Quality

Before fitting any model, run a quick QC summary once. This is especially helpful in multi-omics, clinical, neuroimaging, and socioeconomic workflows where block imbalance or missingness may drive downstream results.

```{r qc}
qc = task_source$overview()
qc$overview
qc$blocks
qc$issues
```

# 5) Define Site Correction

The block assignment already lives on the task, so only the site-correction policy needs to be declared here.
Declared block names are resolved against the columns the MB-sPLS node
receives: a factor declared in a block would claim its encoded dummy columns,
and a column may belong to one block only.

```{r blocks}
site_correction = list(block_a = "site_batch", block_b = "site_batch", block_c = "site_batch")
site_correction_methods = list(block_a = "partial_corr", block_b = "partial_corr", block_c = "partial_corr")

lengths(blocks)
```

# 6) Clone the Clustering Task

The packaged synthetic object is already a `TaskMultiBlock`, so for analysis you can usually just clone it.

```{r task}
task_train = task_source$clone(deep = TRUE)

task_train
length(task_train$feature_names)
```

# 7) Pipelines and Nested CV (Primary Validation)

The analysis compares three pipelines. All share preprocessing, site
correction, and the MB-sPLS fit; they differ only in what reaches the k-means
learner:

- `raw`: the MB-sPLS latent variables (LVs) of the training weights;
- `stable_ci`: LVs recomputed from the features whose bootstrap percentile
  interval excludes zero;
- `stable_frequency`: LVs recomputed from the features selected in at least
  `frequency_threshold` of the bootstrap replicates.

The MB-sPLS node always emits the LVs of its training weights.
`predict_weights` only selects the weights evaluated in its prediction payload,
which the measures and `mbspls_eval_new_data()` read; each pipeline evaluates
the weights its own LVs use. In the stable pipelines,
`PipeOpMBsPLSBootstrapSelect` replaces the raw LVs by the stable LVs at
training and at prediction. It aligns replicate signs separately for every
block and, if the task had an mlr3 `group` role (for example repeated visits
of one participant), would resample whole groups. The explicit
`seed_bootstrap` gives every replicate its own RNG stream; the default `NULL`
uses the session RNG, which `set.seed()` makes reproducible.

```{r pipelines}
modes = list(
  raw = list(selection = FALSE, selection_method = "ci", predict_weights = "raw"),
  stable_ci = list(selection = TRUE, selection_method = "ci", predict_weights = "stable_ci"),
  stable_frequency = list(
    selection = TRUE,
    selection_method = "frequency",
    predict_weights = "stable_frequency"
  )
)

make_mode_learner = function(mode, c_matrix = NULL) {
  mode_spec = modes[[mode]]
  ppl(
    "mbspls_graph_learner",
    learner = lrn("clust.kmeans", centers = cfg$centers),
    task = task_train,
    site_correction = site_correction,
    site_correction_methods = site_correction_methods,
    ncomp = cfg$ncomp,
    c_matrix = c_matrix,
    performance_metric = cfg$perf_metric,
    permutation_test = FALSE,
    predict_weights = mode_spec$predict_weights,
    val_test = "none",
    bootstrap = mode_spec$selection,
    store_train_blocks = TRUE,
    bootstrap_selection = mode_spec$selection,
    selection_method = mode_spec$selection_method,
    frequency_threshold = cfg$frequency_threshold,
    B = cfg$bootstrap_B,
    seed_bootstrap = cfg$seed_bootstrap
  )
}
```

Nested CV estimates held-out performance; it does not by itself perform a
null-hypothesis test. Each outer fold tunes `c_matrix` on its analysis rows
and trains the complete pipeline there, including bootstrap selection for the
stable pipelines, before scoring the assessment rows. Every node with a log
environment gets a fresh one per fold, so the outer estimate covers the
complete configured pipeline. The random search draws its candidates from the
session RNG, so each nested CV starts from the same seed; with the
deterministic MB-sPLS fit and the fixed `seed_bootstrap`, all pipelines then
tune the same `c_matrix` in each fold and differ only in the selection step.

```{r nested-cv}
rs_outer = rsmp("cv", folds = cfg$outer_folds)
rs_inner = rsmp("cv", folds = cfg$inner_folds)
rs_outer$instantiate(task_train)

nested_cv = list()
for (mode in names(modes)) {
  set.seed(cfg$seed_tuning)
  nested_cv[[mode]] = mbspls_nested_cv(
    task = task_train,
    graphlearner = make_mode_learner(mode),
    rs_outer = rs_outer,
    rs_inner = rs_inner,
    ncomp = cfg$ncomp,
    tuner_budget = cfg$tuner_budget,
    tuning_early_stop = FALSE,
    performance_metric = cfg$perf_metric,
    val_test = "none",
    store_payload = TRUE
  )
}

# Same seed, deterministic fit: every pipeline tuned the same c_matrix per fold.
stopifnot(
  identical(nested_cv$raw$c_mats, nested_cv$stable_ci$c_mats),
  identical(nested_cv$raw$c_mats, nested_cv$stable_frequency$c_mats)
)

nested_results = rbindlist(lapply(names(nested_cv), function(mode) {
  result = nested_cv[[mode]]
  result$results[, .(
    mode = mode,
    split,
    weights_evaluated = vapply(result$payloads, `[[`, character(1L), "weights_source"),
    measure_test,
    mac_lv1_test,
    measure_test_n_undefined
  )]
}))
nested_results

nested_summary = rbindlist(lapply(names(nested_cv), function(mode) {
  nested_cv[[mode]]$summary_table[, .(mode = mode, metric, mean, sd, n_defined)]
}))
nested_summary
```

`measure_test` is the held-out EV-weighted MAC (`mbspls.mac_evwt`) of the
weights each pipeline evaluates. `measure_test_n_undefined` counts components
whose held-out association is undefined, for example because stability
selection left fewer than two blocks; such components score zero. With two
outer folds on 90 rows these estimates are very noisy, and choosing the best
pipeline from them would be a further selection step that this nested CV does
not cover. Use more outer folds or repeats, and pre-specify the pipeline, in a
real analysis.

# 8) Retune c_matrix on All Rows

After nested validation, retune on the full training task to obtain a single final sparsity pattern (`c_matrix_final`).
This matrix is then reused across all final pipelines for comparability.

`TunerSeqMBsPLS` resolves the same block columns as the final `PipeOpMBsPLS`,
centres every inner training fold with its own means, and scores the fold
payloads itself, so this `ti()` instance does not need stored models. When the
package measures are scored through `resample()`, `benchmark()`, or `tune()`,
set `store_models = TRUE`. `early_stopping = TRUE` would add a
permutation-based stopping heuristic; it reuses the inner folds that selected
`c`, so it is optimistic, and its `perm_alpha` is a cutoff, not an error rate.

```{r tune}
gl_tune = make_mode_learner("raw")

tuner = TunerSeqMBsPLS$new(
  tuner = "random_search",
  budget = cfg$tuner_budget,
  resampling = rsmp("cv", folds = cfg$inner_folds),
  parallel = "none",
  early_stopping = FALSE,
  performance_metric = cfg$perf_metric
)

instance = ti(
  task = task_train,
  learner = gl_tune,
  resampling = rsmp("insample"),
  measure = msr("mbspls.mac_evwt"),
  terminator = trm("evals", n_evals = 1)
)

set.seed(cfg$seed_tuning)
tuner$optimize(instance)
c_matrix_final = instance$result$learner_param_vals[[1]]$c_matrix
c_matrix_final
instance$result_y
```

At this point, `c_matrix_final` is the most important object to inspect.
It controls feature sparsity per block and per component in the final fits.
`instance$result_y` is the inner-resampling score of the selected candidate.
It was maximised on the same folds and is therefore optimistic; the nested CV
above is the performance estimate.

# 9) Final Fits and What the Selection Modes Change

The three final pipelines use `c_matrix_final`, the same bootstrap seed, and
the same k-means seed. The raw pipeline passes the training LVs to k-means;
the two stable pipelines pass the stable LVs of their selection mode.

```{r final-fits}
final_models = list()
final_summary = list()
final_clusters = list()

for (mode in names(modes)) {
  gl_final = make_mode_learner(mode, c_matrix = c_matrix_final)
  set.seed(cfg$seed_final)
  gl_final$train(task_train)
  prediction = gl_final$predict(task_train)

  # Payload of the weights this pipeline evaluates.
  evaluated = mbspls_eval_new_data(gl_final, task_train)
  n_selected = vapply(evaluated$weights, function(component) {
    sum(vapply(component, function(w) sum(w != 0), integer(1L)))
  }, integer(1L))
  blocks_kept = vapply(evaluated$weights, function(component) {
    kept = vapply(component, function(w) any(w != 0), logical(1L))
    paste(names(component)[kept], collapse = ", ")
  }, character(1L))

  final_summary[[mode]] = data.table(
    mode = mode,
    component = names(evaluated$weights),
    raw_objective = as.numeric(gl_final$model$mbspls$obj_vec),
    weights_evaluated = evaluated$weights_source,
    n_selected = n_selected,
    blocks_kept = blocks_kept,
    train_mac = as.numeric(evaluated$mac_comp),
    train_ev = as.numeric(evaluated$ev_comp)
  )
  final_clusters[[mode]] = as.character(prediction$partition)
  final_models[[mode]] = gl_final
}

# One shared MB-sPLS fit: the raw objectives agree across pipelines.
stopifnot(
  identical(final_models$raw$model$mbspls$obj_vec, final_models$stable_ci$model$mbspls$obj_vec),
  identical(final_models$raw$model$mbspls$obj_vec, final_models$stable_frequency$model$mbspls$obj_vec)
)

final_summary = rbindlist(final_summary)
final_summary

lapply(final_models, function(gl) gl$model$clust.kmeans$train_task$feature_names)
table(raw = final_clusters$raw, stable_ci = final_clusters$stable_ci)
table(raw = final_clusters$raw, stable_frequency = final_clusters$stable_frequency)
```

All pipelines share one MB-sPLS fit, so `raw_objective` is identical by
construction. The mode-specific columns describe the weights each pipeline
evaluates: how many features and which blocks the stability filter kept, and
the resulting training-row MAC and explained variance. A `NaN` MAC means that
fewer than two blocks were kept for that component, so no cross-block
association is defined. When the filter keeps every feature with a non-zero
training weight, as a lenient frequency threshold can, the stable pipeline
reproduces the raw one. These in-sample values are optimistic descriptions,
not performance estimates. The LV features and cross-tabulations show how the
filter changes what k-means sees and how the clustering changes. An honest
comparison of the pipelines requires the selection to be repeated inside
resampling, as in the nested CV of section 7.

# 10) Visualize Validation and Stability

The outer-fold plot shows assessment performance only. It is not a permutation
test and is not interchangeable with the null-distribution plots above.

```{r nested-cv-plot, fig.height=4.5, fig.width=7}
nested_plot_data = nested_results[
  , .(mode, split = factor(split), assessment_score = measure_test)
]
nested_plot = ggplot(
  nested_plot_data,
  aes(x = split, y = assessment_score, fill = mode)
) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("#5B9BD5", "#70AD47", "#ED7D31")) +
  labs(
    x = "Outer fold",
    y = "EV-weighted MAC",
    fill = "Pipeline",
    title = "Nested-CV assessment performance",
    subtitle = "Each outer fold reruns tuning and bootstrap selection"
  ) +
  theme_minimal(base_size = 11)
nested_plot
```

The next plot shows the bootstrap means, percentile intervals, and
selection-frequency opacity of the `stable_ci` pipeline. Replicate weights are
sign-aligned separately for every block. These are stability summaries rather
than feature-level significance tests.

```{r weight-stability-plot, fig.height=7.5, fig.width=12}
weight_stability_plot = mbspls_plot_block_weight_ci(
  final_models$stable_ci,
  source = "bootstrap",
  ci_filter = "none",
  top_n = 4L,
  add_block_rule = FALSE,
  alpha_by_stability = TRUE
)
weight_stability_plot
```

Explained variance provides a complementary component-level description of the
same final model.

```{r variance-plot, fig.height=4.5, fig.width=7}
variance_plot = autoplot(
  final_models$stable_ci,
  type = "mbspls_variance",
  source = "bootstrap",
  show_total = TRUE
)
variance_plot
```

# 11) Inspect Results in Memory

Nested-CV results, fitted models, summaries, predictions, and runtime versions
remain available in the current R session.

`mbspls_model_summary()` extracts reporting tables from fitted MB-sPLS,
MB-sPLS-XY, or MB-sPCA models. Its `conditional_p_value` column holds the
train-time permutation diagnostic (`permutation_test = TRUE`), which conditions
on the fitted preprocessing and hyperparameters; `p_value_scope` states that
scope. Both are `NA` here because the diagnostic was not run.

```{r inspect-results}
names(final_models)
mbspls_model_summary(final_models$raw)$components
final_models$raw$model$mbspls$converged
head(data.table(row_id = task_train$row_ids, cluster = final_clusters$raw))

runtime_versions = data.table(
  package = c("R", "mlr3", "mlr3mbspls"),
  version = c(
    as.character(getRversion()),
    as.character(packageVersion("mlr3")),
    as.character(packageVersion("mlr3mbspls"))
  )
)
runtime_versions
```

# 12) Objects to Inspect

At any point, inspect objects such as:

- `dt_demo`
- `blocks`
- `mb_test_results`
- `complete_test`
- `lc_confirmation$results`
- `lc_confirmation$pairwise_correlations`
- `bootstrap_summary`
- `task_train`
- `nested_results`
- `nested_summary`
- `c_matrix_final`
- `final_summary`
- `final_clusters`
- `final_models$raw$model$mbspls`
- `final_models$stable_ci$model$mbspls_bootstrap_select`
